00-Data-Processing-for-sobiEquity

### 2021-03-13 sobiEquity/raw-data-processing/00-Data-Processing-for-sobiEquity.Rmd
Major reorganization of repository to clean up files and to create (mostly data) package.

---
title: "00-Data-Processing-Example"
output:
  pdf_document: default
  html_notebook: default
---

@@

Data used in this notebook was retrieved from two sources: Open Hamilton and the 2016 Canadian Census. Open Hamilton is an online repository of data from the City of Hamilton. The following datasets were retrieved:

## Provenance of data

### SoBi Hub

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

### Places of Workshop

@@

@@

@@

@@

@@

@@

@@

@@

Load packages used in the notebook:
```{r message = FALSE}
library(disk.frame)
library(pycno)
library(readr)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
library(sf)
library(tidyverse)
```

@@

Load data files:
```{r}
load("hamilton_cma.RData")
load("hamilton_da_2016.RData")
```

@@

Read Sobi service area:
```{r}
sobi_service <- st_read("SoBi_Service_Areas.shp")
```

@@

Read parks:
```{r}
parks <- st_read("Parks.shp")
```

Read cemeteries:
```{r}
cemeteries <- st_read("Cemeteries.shp")
```

Read environmentally sensitive areas:
```{r}
esa <- st_read("Environmentally_Sensitive_Areas_Boundaries.shp")
```

Read golf courses:
```{r}
golf_courses <- st_read("Golf_Courses.shp")
```

Read employment land areas, which include business parks and industrial lands:
```{r}
employment_lands <- st_read("Employment_Lands.shp")
```

Read railways:
```{r}
railways <- st_read("Railways.shp")
```

Read municipal parking lots:
```{r}
municipal_parking <- st_read("Municipal_Parking_Lots.shp")
```

Read street network:
```{r}
streets <- st_read("Street_Centreline.shp")
```

@@

Read educational institutions:
```{r}
schools <- st_read("Educational_Institutions.shp")
```

Read places of worship:
```{r}
religious_places <- st_read("Places_of_Worship.shp")
```

Read municipal service centres:
```{r}
service_centres <- st_read("Municipal_Service_Centres.shp")
```

Read recreation and centres:
```{r}
recreation_centres <- st_read("Recreation_and_Community_Centres.shp")
```

Read arenas:
```{r}
arenas <- st_read("Arenas.shp")
```

Read EMS stations:
```{r}
ems <- st_read("EMS_Stations.shp")
```

Read fire stations:
```{r}
fire <- st_read("Fire_Stations.shp")
```

Read police stations:
```{r}
police <- st_read("Police_Stations.shp")
```

Read hospitals:
```{r}
hospitals <- st_read("Hospitals.shp")
```

@@

@@

Extract railways in service area:
```{r}
railways_sobi <- st_crop(railways, st_as_sfc(bounding_box))
```
Extract municipal parking lots in service area:
```{r}
parking_sobi <- st_crop(municipal_parking, st_as_sfc(bounding_box))
```

@@

### Read population

Read population table (2016 Census of Population, Population and Dwelling counts):
```{r}
population_da_2016 <- read_csv("T1901EN.csv") %>%
  transmute(DA = factor(`Geographic code`), population = `Population, 2016`)
```

@@

Remove EMS stations from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, ems_sobi)
```
Remove fire stations from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, fire_sobi)
```
Remove police stations from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, police_sobi)
```
Remove hospitals from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, hospitals_sobi)
```

@@

Read Sobi hubs:
```{r}
sobi_hubs <- st_read("SoBi_Hubs.shp")
```

@@

Plot again:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop_clean,
          aes(color = population)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = sobi_hubs,
          aes(size = RACKS_AMOU)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  scale_size(range = c(1, 4))
```

Remove ERI hubs that were implemented in 2018 to increase accessibility for under-serviced and disadvantaged areas of the city. This reflects the original state of the system without the equity stations:
```{r}
sobi_hubs_original <- sobi_hubs %>%
  dplyr::filter(NAME != "Mars at Wentworth - ERI01") %>%
  dplyr::filter(NAME != "Sherman at Barton - ERI02") %>%
  dplyr::filter(NAME != "Barton at Lottridge - ERI03") %>%
  dplyr::filter(NAME != "Barton and Belview - ERI05") %>%
  dplyr::filter(NAME != "Barton at Ottawa - ERI06") %>%
  dplyr::filter(NAME != "Ottawa at Dunsmure - ERI07") %>%
  dplyr::filter(NAME != "Maple at Rothsay - ERI08") %>%
  dplyr::filter(NAME != "King at Dunsmure - ERI09") %>%
  dplyr::filter(NAME != "Dunsmure at Sherman - ERI10") %>%
  dplyr::filter(NAME != "Gage at Cannon - ERI11") %>%
  dplyr::filter(NAME != "Belview at Cannon - ERI12") %>%
  dplyr::filter(NAME != "Westinghouse at Barton - ERI13")
```

Plot again:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop_clean,
          aes(color = population)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = sobi_hubs_original,
          aes(size = RACKS_AMOU)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  scale_size(range = c(1, 4))
```
After removing the twelve ERI hubs, there is much less coverage in the east end of the service area. Again, the service area is the area where a cyclist can use a SoBi bike. It makes sense then that ERI would want to fill this gap in coverage because individuals who live in the east end of the service area could theoretically use the bike share system but may not live near a hub. For this reason, twelve hubs were added in 2018 to expand the system.

@@

## Routing for original system without ERI hubs

Build the graph:
```{r build-graph-original, include = FALSE}
r5_sobi_original <- setup_r5(data_path = r5_path, verbose = FALSE)
```

Prepare Input Data for `r5r`. The origins are the coordinates of the population cells and the destinations the coordinates of the SoBi hubs (minus the ERI hubs in the east end of the service area):
```{r prepare-inputs-original}
# save origin centroids in format expected by R5R (id, lon, lat)
origins_io <- data.frame(UID = interpolated_pop_clean$UID, 
              interpolated_pop_clean %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = UID) %>%
    dplyr::select(id, lon, lat)

# now SoBi hubs
destinations_jo <- data.frame(OBJECTID = sobi_hubs_original$OBJECTID, 
              sobi_hubs_original %>%
                st_transform(crs = 4326) %>%
                st_coordinates()) %>%
    rename(lon = X, lat = Y, id = OBJECTID) %>%
    dplyr::select(id, lon, lat)
```

Calculate OD Matrix for Walking:
```{r calculate walking od matrix original system, include=FALSE}
# set up batching according to how many origin rows to process at one time
chunksize_o = 2000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
num_chunks_o = ceiling(nrow(origins_io)/chunksize_o)

# create origin-destination pairs
origins_chunks_o <- as.disk.frame(origins_io,
                          outdir = "./df/origins_io",
                          nchunks = num_chunks_o,
                          overwrite = TRUE)

start.time_o <- Sys.time()
pb_o <- txtProgressBar(0, num_chunks_o, style = 3)

for (io in 1:num_chunks_o){
  origins_io_chunk <- get_chunk(origins_chunks_o, io)
  ttm_chunk_o <- travel_time_matrix(r5_sobi_original,
                          origins = origins_io_chunk,
                          destinations = destinations_jo,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2020-03-02 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_walk_dist = 10000, # metres
                          max_trip_duration = 30)

  # export output as disk.frame
  ifelse(io == 1, output_o_df <- as.disk.frame(ttm_chunk_o,
                                            nchunks = 1,
                                            outdir = "./df/output_ttm_walk_o",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_o_df, ttm_chunk_o, chunk_id = io))
  setTxtProgressBar(pb_o, io)
}
end.time_o <- Sys.time()
ttm_walk_time_o <- end.time - start.time
ttm_walk_time_o
```

## Extract travel time matrix for original system

```{r load od matrix for walk and sobi, include=FALSE}
# connect to the walking travel time matrix disk frame
ttm_walk_o.disk.frame <- disk.frame("./df/output_ttm_walk_o")
#opportunities_j <- as.data.frame(nyc_cb_point) %>% select(GEOID10, total_emp) %>% rename(toId = GEOID10, o_j = total_emp)
```

Convert disk.frame to data frame:
```{r}
ttm_walk_o <- as.data.frame(ttm_walk_o.disk.frame) %>%
  transmute(UID = as.numeric(fromId), OBJECTID = as.numeric(toId), travel_time)
```

@@

Rename data frame:
```{r}
population_sobi <- interpolated_pop_clean
```

Save data files:
```{r}
save(population_sobi, file = "population_sobi.RData")
save(sobi_hubs, file = "sobi_hubs.RData")
save(sobi_hubs_original, file = "sobi_hubs_original.RData")
save(sobi_service, file = "sobi_service.RData")
save(ttm_walk, file = "ttm_walk.RData")
save(ttm_walk_o, file = "ttm_walk_original.RData")
save(hamilton_da_sobi, file = "hamilton_da_sobi.RData")
save(interpolated_pop_clean, file = "interpolated_pop_clean.RData")
```

>>>>

---
title: "00-Data-Processing-for-sobiEquity"
---

@@

Load packages used in the notebook:
```{r message = FALSE}
library(cancensus)
library(disk.frame)
library(pycno)
library(readr)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
library(sf)
library(tidyverse)
```

Data used in this notebook was retrieved from two sources: Open Hamilton and the 2016 Canadian Census. Open Hamilton is an online repository of data from the City of Hamilton. The provenance of the datasets is below.

## Provenance of data: City of Hamilton

Source?

### SoBi Hubs

`SoBi_Hubs.shp`

@@

`SoBi_Service_Areas.shp`

@@

`Golf_Courses.shp`

@@

`Parks.shp`

@@

`Employment_Lands.shp`

@@

`Municipal_Parking_Lots.shp`

@@

`Cemeteries.shp`

@@

`Environmentally_Sensitive_Areas_Boundaries.shp`

@@

`Street_Centreline.shp`

@@

`Educational_Institutions.shp`

@@

### Places of Worship

`Places_of_Worship.shp`

@@

`Municipal_Service_Centres.shp`

@@

`Recreation_and_Community_Centres.shp`

@@

`Arenas.shp`

@@

`EMS_Stations.shp`

@@

`Fire_Stations.shp`

@@

`Police_Stations.shp`

@@

`Hospitals.shp`

@@

### Railways

## Provenance of data: census

We can get an API key to use `cancensus` from [CensusMapper](https://censusmapper.ca/). Once you have an API key, it can be stored locally this way:
```{r eval=FALSE}
set_api_key("your-key", install = TRUE)
```

Once you have your API you can use `cancensus` like in this example.

Use `cancensus` to check and download census data. Check the regions:
```{r}
list_census_regions('CA16') %>% 
  dplyr::filter(level == "CMA", name %in% c("Hamilton"))
```

We need PR_UID 35.

List census data sets:
```{r}
list_census_datasets()
```

We need CA16.

Search variables:
```{r}
find_census_vectors("population", 
                    dataset = "CA16", 
                    type = "total", 
                    query_type = "keyword", 
                    interactive = FALSE)
```

```{r}
find_census_vectors("median household income", 
                    dataset = "CA16", 
                    type = "total", 
                    query_type = "keyword", 
                    interactive = FALSE)
```

We need v_CA16_401 and v_CA16_2397.

Retrieve data:
```{r}
population_da_2016 <- get_census(dataset='CA16', 
                          regions=list(CMA=c("35537")),
                          vectors=c("v_CA16_401"),
                          level='DA', 
                          use_cache = FALSE)

mhi <- get_census(dataset='CA16', 
                          regions=list(CMA=c("35537")),
                          vectors=c("v_CA16_2397"),
                          level='DA', 
                          use_cache = FALSE)
```

@@

Load data files:
```{r}
load("input-data-files/hamilton_cma.RData")
load("input-data-files/hamilton_da_2016.RData")
```

@@

Read Sobi service area:
```{r}
sobi_service <- st_read("input-data-files/SoBi_Service_Areas.shp")
```

@@

Read parks:
```{r}
parks <- st_read("input-data-files/Parks.shp")
```

Read cemeteries:
```{r}
cemeteries <- st_read("input-data-files/Cemeteries.shp")
```

Read environmentally sensitive areas:
```{r}
esa <- st_read("input-data-files/Environmentally_Sensitive_Areas_Boundaries.shp")
```

Read golf courses:
```{r}
golf_courses <- st_read("input-data-files/Golf_Courses.shp")
```

Read employment land areas, which include business parks and industrial lands:
```{r}
employment_lands <- st_read("input-data-files/Employment_Lands.shp")
```

Read railways:
```{r}
railways <- st_read("input-data-files/Railways.shp")
```

Read municipal parking lots:
```{r}
municipal_parking <- st_read("input-data-files/Municipal_Parking_Lots.shp")
```

Read street network:
```{r}
streets <- st_read("input-data-files/Street_Centreline.shp")
```

@@

Read educational institutions:
```{r}
schools <- st_read("input-data-files/Educational_Institutions.shp")
```

Read places of worship:
```{r}
religious_places <- st_read("input-data-files/Places_of_Worship.shp")
```

Read municipal service centres:
```{r}
service_centres <- st_read("input-data-files/Municipal_Service_Centres.shp")
```

Read recreation and centres:
```{r}
recreation_centres <- st_read("input-data-files/Recreation_and_Community_Centres.shp")
```

Read arenas:
```{r}
arenas <- st_read("input-data-files/Arenas.shp")
```

Read EMS stations:
```{r}
ems <- st_read("input-data-files/EMS_Stations.shp")
```

Read fire stations:
```{r}
fire <- st_read("input-data-files/Fire_Stations.shp")
```

Read police stations:
```{r}
police <- st_read("input-data-files/Police_Stations.shp")
```

Read hospitals:
```{r}
hospitals <- st_read("input-data-files/Hospitals.shp")
```

@@

Recreate the bounding box:
```{r}
bounding_box <- st_bbox(hamilton_da_sobi)
```

@@

Extract railways in service area:
```{r}
railways_sobi <- st_crop(railways, st_as_sfc(bounding_box))
```

Extract municipal parking lots in service area:
```{r}
parking_sobi <- st_crop(municipal_parking, st_as_sfc(bounding_box))
```

@@

### Prepare population data

Transmute to retain only variables of interest:
```{r}
population_da_2016 <- population_da_2016 %>%
  transmute(DA = factor(`GeoUID`), population = Population)
```

@@

Remove EMS stations from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, ems_sobi)
```
Remove fire stations from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, fire_sobi)
```
Remove police stations from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, police_sobi)
```

Remove hospitals from the DAs (use `st_difference()`):
```{r}
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, hospitals_sobi)
```

@@

Read Sobi hubs:
```{r}
sobi_hubs <- st_read("input-data-files/SoBi_Hubs.shp")
```

@@

Plot again:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop_clean,
          aes(color = population)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = sobi_hubs,
          aes(shape = hub_type, 
              size = RACKS_AMOU)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  scale_size(range = c(1, 4))
```

Identify ERI hubs that were implemented in 2018 to increase accessibility for under-serviced and disadvantaged areas of the city. This reflects the original state of the system without the equity stations:
```{r}
sobi_hubs <- sobi_hubs %>%
  dplyr::mutate(hub_type = case_when(NAME == "Mars at Wentworth - ERI01" ~ "ERI",
                                     NAME == "Sherman at Barton - ERI02" ~ "ERI",
                                     NAME == "Barton at Lottridge - ERI03" ~ "ERI",
                                     NAME == "Barton and Belview - ERI05" ~ "ERI",
                                     NAME == "Barton at Ottawa - ERI06" ~ "ERI",
                                     NAME == "Ottawa at Dunsmure - ERI07" ~ "ERI",
                                     NAME == "Maple at Rothsay - ERI08" ~ "ERI",
                                     NAME == "King at Dunsmure - ERI09" ~ "ERI",
                                     NAME == "Dunsmure at Sherman - ERI10" ~ "ERI",
                                     NAME == "Gage at Cannon - ERI11" ~ "ERI",
                                     NAME == "Belview at Cannon - ERI12" ~ "ERI",
                                     NAME == "Westinghouse at Barton - ERI13" ~ "ERI",
                                     TRUE ~ "Conventional"),
                hub_type = factor(hub_type))

```

Plot again:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop_clean,
          aes(color = population)) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = sobi_hubs,
          aes(shape = hub_type,
              size = RACKS_AMOU)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  scale_size(range = c(1, 4)) + 
  facet_wrap(~hub_type)
```

After removing the twelve ERI hubs, there is much less coverage in the east end of the service area. Again, the service area is the area where a cyclist can use a SoBi bike. It makes sense then that ERI would want to fill this gap in coverage because individuals who live in the east end of the service area could theoretically use the bike share system but may not live near a hub. For this reason, twelve hubs were added in 2018 to expand the system.

@@

Join hub type information to travel time table:
```{r}
ttm_walk <- ttm_walk %>%
  left_join(sobi_hubs %>%
              st_drop_geometry() %>%
              dplyr::select(OBJECTID,
                            hub_type),
            by = "OBJECTID")
```

@@

Rename data frame:
```{r}
population_50x50 <- interpolated_pop_clean
```

Clean DA boundaries:
```{r}
hamilton_da_2016 <- hamilton_da_2016 %>%
  dplyr::select(-POP2011) %>%
  left_join(population_da_2016,
            by = "DA") %>%
  left_join(mhi %>%
              transmute(GeoUID, median_total_income = `v_CA16_2397: Median total income of households in 2015 ($)`),
            by = c("DA" = "GeoUID")) %>%
  left_join(hamilton_da_sobi %>%
              st_drop_geometry() %>%
              transmute(DA,
                        sobi_service = "Yes"),
            by = "DA") %>%
  mutate(sobi_service = ifelse(is.na(sobi_service), "No", "Yes") %>%
           factor())
```

Save data files:
```{r}
save(hamilton_cma, file = "output-data-files/hamilton_cma.RData", compress = "xz")
save(hamilton_da_2016, file = "output-data-files/hamilton_da_2016.RData", compress = "xz")
save(population_50x50, file = "output-data-files/population_50x50.RData", compress = "xz")
save(sobi_hubs, file = "output-data-files/sobi_hubs.RData", compress = "xz")
save(sobi_service, file = "output-data-files/sobi_service.RData", compress = "xz")
save(ttm_walk, file = "output-data-files/ttm_walk.RData", compress = "xz")
```

Copy files to folder `data` of the package.
